Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-02-12 paper.tex
Las noch eine wertvolle Bemerkung erwägen

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}.
\begin{figure*}[tp]
  \centering
  \subfigure[An original image of an anisotropic pore.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso.png}
    \label{fig:aniso-unrot}}
  \hfill
  \subfigure[A rotated image of the pore, so it is elongated along the axis
    $X$.]{
    \includegraphics[width=0.4\linewidth, frame]{images/aniso-rot.png}
    \label{fig:aniso-rot}}
  \vskip\baselineskip
  \subfigure[$S_3$ function for the original image]{
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}}
  \hfill
  \subfigure[$S_3$ function for the rotated image. A vector $(0.92, -0.4)$ is a
    direction of elongation of the pore.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}
    \label{fig:aniso-rot-s3}}
  \hfill
  \subfigure[$S_3$ function averaged across 100 random rotations.]{
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
    \label{fig:aniso-avg}}
  \caption[]{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. A result on \cref{fig:aniso-avg} averaged
    across rotations is similar to the algorithm described in \cite{berryman1988,SMITH1988176,malmir2018}. \highlight{Make empty image over (e)}}
  \label{fig:aniso}
\end{figure*}

@@

We have made execuation time comparisons of our algorithm with existing algorithms.
However, it is also beneficial to compare some algorithmic details against existing digital approaches --
proposed by Malmir~et~al \cite{malmir2018} and another one by Smith and
Torquato \cite{SMITH1988176}. The first algorithm generates a triangle with
fixed sides $r_1$ and $r_2$ and an arbitrary angle $\theta$ between them and
places it in the medium firstly making a random translation and a rotation
around a random vector by a random angle. By making a large number of trials the
algorithm gets a value of a three-point correlation function at the point
$(r_1, r_2, \theta)$. The second algorithm takes a precalculated triangle
$(r_1, r_2, \theta)$ from a web-like pattern and places it in the medium without
rotation, only applying a random translation. This way the algorithm preserves
orientation of triangles stored in the pattern. The main advantage of our
approach based on right triangle is in its parallelizability and ease of implementation. A drawback is
that an angle between catheti of the sampling pattern is fixed to be
$\theta = \frac{\pi}{2}$, but can be mitigated by rotation of the pattern as a whole by an
equivalent operation of rotation of a sample. These rotations can be random as
in the Malmir et al. algorithm (\cref{fig:aniso-avg}), or it can be a single rotation
to align catheti of the sampling triangle with the direction of anisotropy
(\cref{fig:aniso-rot-s3}). The latter allows computing directional 3-point correlation functions
for anisotropic structures similar to 2-point counterparts \cite{10.1063/1.4867611,EPL1}.

>>>>

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}.
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

We have made execuation time comparisons of our algorithm with existing algorithms.
However, it is also beneficial to compare some algorithmic details against existing digital approaches --
proposed by Malmir~et~al \cite{malmir2018} and another one by Smith and
Torquato \cite{SMITH1988176}. The first algorithm generates a triangle with
fixed sides $r_1$ and $r_2$ and an arbitrary angle $\theta$ between them and
places it in the medium firstly making a random translation and a rotation
around a random vector by a random angle. By making a large number of trials the
algorithm gets a value of a three-point correlation function at the point
$(r_1, r_2, \theta)$. The second algorithm takes a precalculated triangle
$(r_1, r_2, \theta)$ from a web-like pattern and places it in the medium without
rotation, only applying a random translation. This way the algorithm preserves
orientation of triangles stored in the pattern. The main advantage of our
approach based on right triangle is in its parallelizability and ease of implementation. A drawback is
that an angle between catheti of the sampling pattern is fixed to be
$\theta = \frac{\pi}{2}$, but can be mitigated by rotation of the pattern as a whole by an
equivalent operation of rotation of a sample. These rotations can be random as
in the Malmir et al. algorithm, or it can be a single rotation
to align catheti of the sampling triangle with the direction of anisotropy
(\cref{fig:aniso}). The latter allows computing directional 3-point correlation functions
for anisotropic structures similar to 2-point counterparts \cite{10.1063/1.4867611,EPL1}.
